Macroscopic Salt Rejection through Electrostatically Gated Porous Graphene

### 2019-01-15 paper/paper.tex
fix typo
Remove blank line

We next discuss the effects of electrostatic gating on graphene. In
order to avoid any interference from the conductivity probe to the
membrane potential, the ``three-interval'' method, which has been used
in controlling the membrane potential in mesoporous carbon membranes
\cite{Surwade_2014}, is adopted here. \Fig{} \ref{fig:2}a
illustratively represents how the measured conductivity evolves with
time. Specifically, in the first interval, the gate voltage source is
turned off and the conductivity probe is on, allowing us to obtain the
conductivity-time profile with an average slope $s_{1}$ that
corresponds to the diffusive flux across the PG-PCTE membrane. In the
second interval, a voltage $V_{\mathrm{G}}$ is applied to the membrane
after turning off the conductivity probe, followed by the third
interval $s_{3}$, in which the conductivity probe is turned on again
to give the corresponding slope. A control experiment was performed to
ensure the copper tape was not in direct contact with the ionic
solution (Supplementary Section \ref{SI-sec:copper}). The slope
corresponding to the second interval $s_{2}$, as the conductivity
probe is off, is determined by linear interpolation of the
conductivities from the endpoint of interval 1 to the starting point
of interval 3. A decrease of slope observed in interval 2 indicates
that the total flux through PG-PCTE is reduced upon gating (i.e. salt
penetration through the PG pores is hindered). We define such
experimentally measured reduction rate as $\eta$, which is equivalent
to the decrease of total diffusive flux through PG-PCTE as follows:
\begin{equation}
  \label{eq:rejection}
  \eta = \frac{\bar{s} - s_{2}}{\bar{s}} = \frac{\boldsymbol{J}_{\mathrm{PG0}}
    - \boldsymbol{J}_{\mathrm{PG}}(V_{\mathrm{G}})}{\boldsymbol{J}_{\mathrm{PG0}}}
\end{equation}
where $\bar{s}$ is the average slope of $s_{1}$ and $s_{3}$, namely
$(s_{1} + s_{3})/2$, and
$\boldsymbol{J}_{\mathrm{PG}}(V_{\mathrm{G}})$ is the diffusive flux
across the PTCE-supported PG double layer as a function of
$V_{\mathrm{G}}$. Since graphene and PCTE membranes effectively form a
transport system with series diffusive resistance $R_{\mathrm{G}}$
(tunable by $V_{\mathrm{G}}$) and $R_{\mathrm{PCTE}}$ (independent of
$V_{\mathrm{G}}$), respectively, we define the effective degree of
salt rejection from bare PG $\xi$ as a function of $V_{\mathrm{G}}$ (
details see Supplementary Section \ref{SI-sec:R-model}) as:
\begin{equation}
\label{eq:xi-def}
\xi = 1 - \frac{R_{\mathrm{G}}(V_{\mathrm{G}}=0)}{R_{\mathrm{G}}(V_{\mathrm{G}})} = \frac{(\delta+1) \eta}{\delta \eta + 1}
\end{equation}
The factor $\xi$ is a reliable and stable measure of salt rejection
through the PG membrane itself upon gating, as the salt species and
initial conditions may induce a degree of measurement uncertainty
between different samples. Note that the $\xi=1$ limit represents
perfect salt rejection in which
$R_{\mathrm{G}}(V_{\mathrm{G}}) \to \infty$. On the contrary $xi=0$
indicates $R_{\mathrm{G}}$ remains unchanged with $V_{\mathrm{G}}$.

@@

In order to quantitatively understand the observed $V_{\mathrm{G}}$
dependence, we develop a theory to model the coupling of graphene's
elementary transport properties and EDL, in order to quantify the
ionic transport through a graphene nanopore. Under the assumption that
the time scale for the bulk concentration change is significantly
longer than that of ionic diffusion, the pseudo-steady state
approximation holds. Accordingly, the steady-state Nernst-Planck
equation describing the ionic transport in an electrolyte solution is
given by \cite{MacGillivray_1968}:
\begin{equation}
  \label{eq:pnp}
  \nabla \cdot \boldsymbol{J}_{i} = -\nabla \cdot (\frac{D_{i}}{k_{\mathrm{B}}T} c_{i} N_{\mathrm{A}} \nabla \mu_{i}) = 0
\end{equation}
where $\boldsymbol{J}$ is the mass flux, subscript \textit{i}
corresponds to the ionic species (anion or cation), $D$ is the diffusivity,
$k_{\mathrm{B}}$ is the Boltzmann constant, $T$ is temperature, $c$ is
the molar concentration, $N_{\mathrm{A}}$ is the Avogadro constant,
and $\mu_{i}$ is the electrochemical potential. Under the assumption
of ideal solution, it follows \cite{Kilic_2007}:
\begin{equation}
  \label{eq:mu}
  \nabla \mu_{i} = k_{\mathrm{B}} T \nabla \ln x_{i} + z_{i} e \nabla \psi
\end{equation}
 $x$ is the molar fraction in solution, $z$ is the ionic valence,
$e$ is the unit charge, and $\psi$ is the electric potential. On the
other hand, the Poisson equation describing the electric potential
distribution is given by:
\begin{equation}
  \label{eq:poisson}
  \nabla \cdot (\varepsilon_{\mathrm{m}} \varepsilon_{0} \nabla \psi)
  =
  - N_{\mathrm{A}} e \sum_{i} c_{i} z_{i}
\end{equation}
where $\varepsilon_{\mathrm{m}}$ is the relative permittivity of
individual materials in the system, including water and the internal
Stern layer\cite{Tian_2017}, and
$\varepsilon_{0}$ is the vacuum permittivity. When applying
$V_{\mathrm{G}}$ to graphene adjacent to the electrolyte solution,
charges (electrons or holes) are induced in graphene; the
electroneutrality of the entire system suggests:
\begin{equation}
  \label{eq:electro-neutral}
  \sigma_{\mathrm{G}} S_{\mathrm{G}} + \sum_{i} \int_{\Omega} z_{i} c_{i} N_{\mathrm{A}} e \mathrm{d}^{3} \Omega= 0
\end{equation}
where $\sigma_{\mathrm{G}}$ is the charge density in graphene,
$S_{\mathrm{G}}$ is the total area of graphene, and $\Omega$
corresponds to the entire volumetric domain of electrolyte
solution. Note that the graphene surface potential,
$\psi_{\mathrm{G}}$, is not equivalent to $V_{\mathrm{G}}$ due to
change of graphene's work function upon charging, known as the
graphene quantum capacitance effect\cite{Xia_2009} (\Fig{}
\ref{fig:3}a) following:
\begin{equation}
  \label{eq:Vg}
  V_{\mathrm{G}} = \Delta \phi_{\mathrm{G}} + \psi_{\mathrm{G}}
\end{equation}
where
$\Delta \phi_{\mathrm{G}} = \phi_{\mathrm{G}}(V_{\mathrm{g}}) -
\phi_{\mathrm{G}}(V_{\mathrm{g}}=0)$ is the change of graphene’s work
function. The properties of the PG samples fabricated is close to
intrinsic graphene, with charge density at $V_{\mathrm{G}}=0$ measured
to be 1.4$\times$10$^{11}$ \textit{e}$\cdot$cm$^{-2}$, corresponding
to a work function difference of 48 meV from graphene's Dirac point
(details see Supplementary Section \ref{SI-sec:charge-dens}), much
smaller than the $\Delta \phi_{\mathrm{G}}$ induced by gating.
Therefore we adapt the simplification that the charge neutrality point
of graphene coincides with the Dirac point at $V_{\mathrm{G}}$ =
0. The elementary electronic properties of graphene give:

\begin{equation}
  \label{eq:delta-phiG}
  \Delta \phi_{\mathrm{G}} = \mathrm{sign}(\sigma_{\mathrm{G}}) \frac{\hbar v_{\mathrm{F}}}{e}
  \sqrt{\frac{\pi |\sigma_{\mathrm{G}}|}{e}}
\end{equation}

where $\hbar$ is the reduced Planck constant, and
$v_{\mathrm{F}}=1.1\times10^{6}$ m$\cdot$s$^{-1}$ is the Fermi
velocity of graphene (more details find Supplementary Section
\ref{SI-sec:charge-dens}).  Note that the PG film fabricated
experimentally is double-layer turbostratic (randomly aligned)
graphene, in which we assume the
$\Delta \phi_{\mathrm{G}} - \sigma_{\mathrm{G}}$ dependence follows
Eq. \eqref{eq:delta-phiG}. The theory proposed here was solved
self-consistently by discretizing
Eqs. \eqref{eq:pnp}-\eqref{eq:poisson} using the finite-element method
(FEM), in which the graphene surface potential is coupled with
Eqs. \eqref{eq:electro-neutral}-\eqref{eq:delta-phiG} that were solved
simultaneously to reach convergent numerical solutions of
$\sigma_{\mathrm{G}}$ and $\psi_{\mathrm{G}}$ for a given
$V_{\mathrm{G}}$. Clearly, the above model yields symmetric
characteristics for $\sigma_{\mathrm{G}}$ and $\psi_{\mathrm{G}}$ with
respect to $V_{\mathrm{G}}$ due to the symmetric band structure of
graphene (Eq. \eqref{eq:delta-phiG}). However, for graphene
synthesized by chemical vapor deposition (CVD) used in our
experiments, it is well-recognized that the electron traps are
inherently introduced during synthesis and patterning
\cite{Dean_2010}, effectively reducing the charge density and surface
potential on graphene $V_{\mathrm{G}}<0$ (Supplementary \Fig{}
\ref{SI-fig:trap}), which is related to the fact that observed $\xi$
is much lower within the $V_{\mathrm{G}}<0$ regime in \Fig{}
\ref{fig:2}c. With the nonideality in mind, hereafter, we compare the
calculations and experiments in the regime of $V_{\mathrm{G}}>0$. For
example, \Fig{} \ref{fig:3}c presents the calculated
$\sigma_{\mathrm{G}}$ and $\psi_{\mathrm{G}}$ as a function of
$V_{\mathrm{G}}$ considering the ionic diffusion through a single 20
nm-diameter nanopore drilled on a sheet of semi-infinite double layer
graphene that separates HCR containing KCl solution at $c_{0}$ = 0.1
mM, and LCR at $c_{0}/10$ in axisymmetric coordinates (details see
Supplementary Section \ref{SI-fig:scheme}). Indeed, the interplay
between graphene quantum capacitance and the EDL significantly reduces
the graphene surface potential $\psi_{\mathrm{G}}$ to $\sim$0.3 V at
$V_{\mathrm{G}}$ = 1.25 V, corresponding a surface charge density
$\sigma_{\mathrm{G}}$ of $\sim$0.08 C$\cdot$m$^{-2}$. Note that as we
mentioned earlier, an important merit of the 2D nanopore system
considered here is that the surface charge density can be precisely
determined and controlled, rather than being treated as a fitting
parameter as in most of the 2D nanopore literature
(e.g. Ref. \citenum{Rollings_2016}). In addition, we anticipate that
the gate-controlled ionic transport can also be achieved by replacing
graphene with other nanoporous conducting 2D materials, although the
operational gate voltage window required to achieve the same degree of
salt rejection may be narrower, because graphene has the smallest
quantum capacitance (near the Dirac point) among all known 2D
materials \cite{Tian_2016}.

>>>>

We next discuss the effects of electrostatic gating on graphene. In
order to avoid any interference from the conductivity probe to the
membrane potential, the ``three-interval'' method, which has been used
in controlling the membrane potential in mesoporous carbon membranes
\cite{Surwade_2014}, is adopted here. \Fig{} \ref{fig:2}a
illustratively represents how the measured conductivity evolves with
time. Specifically, in the first interval, the gate voltage source is
turned off and the conductivity probe is on, allowing us to obtain the
conductivity-time profile with an average slope $s_{1}$ that
corresponds to the diffusive flux across the PG-PCTE membrane. In the
second interval, a voltage $V_{\mathrm{G}}$ is applied to the membrane
after turning off the conductivity probe, followed by the third
interval $s_{3}$, in which the conductivity probe is turned on again
to give the corresponding slope. A control experiment was performed to
ensure the copper tape was not in direct contact with the ionic
solution (Supplementary Section \ref{SI-sec:copper}). The slope
corresponding to the second interval $s_{2}$, as the conductivity
probe is off, is determined by linear interpolation of the
conductivities from the endpoint of interval 1 to the starting point
of interval 3. A decrease of slope observed in interval 2 indicates
that the total flux through PG-PCTE is reduced upon gating (i.e. salt
penetration through the PG pores is hindered). We define such
experimentally measured reduction rate as $\eta$, which is equivalent
to the decrease of total diffusive flux through PG-PCTE as follows:
\begin{equation}
  \label{eq:rejection}
  \eta = \frac{\bar{s} - s_{2}}{\bar{s}} = \frac{\boldsymbol{J}_{\mathrm{PG0}}
    - \boldsymbol{J}_{\mathrm{PG}}(V_{\mathrm{G}})}{\boldsymbol{J}_{\mathrm{PG0}}}
\end{equation}
where $\bar{s}$ is the average slope of $s_{1}$ and $s_{3}$, namely
$(s_{1} + s_{3})/2$, and
$\boldsymbol{J}_{\mathrm{PG}}(V_{\mathrm{G}})$ is the diffusive flux
across the PTCE-supported PG double layer as a function of
$V_{\mathrm{G}}$. Since graphene and PCTE membranes effectively form a
transport system with series diffusive resistance $R_{\mathrm{G}}$
(tunable by $V_{\mathrm{G}}$) and $R_{\mathrm{PCTE}}$ (independent of
$V_{\mathrm{G}}$), respectively, we define the effective degree of
salt rejection from bare PG $\xi$ as a function of $V_{\mathrm{G}}$ (
details see Supplementary Section \ref{SI-sec:R-model}) as:
\begin{equation}
\label{eq:xi-def}
\xi = 1 - \frac{R_{\mathrm{G}}(V_{\mathrm{G}}=0)}{R_{\mathrm{G}}(V_{\mathrm{G}})} = \frac{(\delta+1) \eta}{\delta \eta + 1}
\end{equation}
The factor $\xi$ is a reliable and stable measure of salt rejection
through the PG membrane itself upon gating, as the salt species and
initial conditions may induce a degree of measurement uncertainty
between different samples. Note that the $\xi=1$ limit represents
perfect salt rejection in which
$R_{\mathrm{G}}(V_{\mathrm{G}}) \to \infty$. On the contrary $\xi=0$
indicates $R_{\mathrm{G}}$ remains unchanged with $V_{\mathrm{G}}$.

@@

In order to quantitatively understand the observed $V_{\mathrm{G}}$
dependence, we develop a theory to model the coupling of graphene's
elementary transport properties and EDL, in order to quantify the
ionic transport through a graphene nanopore. Under the assumption that
the time scale for the bulk concentration change is significantly
longer than that of ionic diffusion, the pseudo-steady state
approximation holds. Accordingly, the steady-state Nernst-Planck
equation describing the ionic transport in an electrolyte solution is
given by \cite{MacGillivray_1968}:
\begin{equation}
  \label{eq:pnp}
  \nabla \cdot \boldsymbol{J}_{i} = -\nabla \cdot (\frac{D_{i}}{k_{\mathrm{B}}T} c_{i} N_{\mathrm{A}} \nabla \mu_{i}) = 0
\end{equation}
where $\boldsymbol{J}$ is the mass flux, subscript \textit{i}
corresponds to the ionic species (anion or cation), $D$ is the diffusivity,
$k_{\mathrm{B}}$ is the Boltzmann constant, $T$ is temperature, $c$ is
the molar concentration, $N_{\mathrm{A}}$ is the Avogadro constant,
and $\mu_{i}$ is the electrochemical potential. Under the assumption
of ideal solution, it follows \cite{Kilic_2007}:
\begin{equation}
  \label{eq:mu}
  \nabla \mu_{i} = k_{\mathrm{B}} T \nabla \ln x_{i} + z_{i} e \nabla \psi
\end{equation}
 $x$ is the molar fraction in solution, $z$ is the ionic valence,
$e$ is the unit charge, and $\psi$ is the electric potential. On the
other hand, the Poisson equation describing the electric potential
distribution is given by:
\begin{equation}
  \label{eq:poisson}
  \nabla \cdot (\varepsilon_{\mathrm{m}} \varepsilon_{0} \nabla \psi)
  =
  - N_{\mathrm{A}} e \sum_{i} c_{i} z_{i}
\end{equation}
where $\varepsilon_{\mathrm{m}}$ is the relative permittivity of
individual materials in the system, including water and the internal
Stern layer\cite{Tian_2017}, and
$\varepsilon_{0}$ is the vacuum permittivity. When applying
$V_{\mathrm{G}}$ to graphene adjacent to the electrolyte solution,
charges (electrons or holes) are induced in graphene; the
electroneutrality of the entire system suggests:
\begin{equation}
  \label{eq:electro-neutral}
  \sigma_{\mathrm{G}} S_{\mathrm{G}} + \sum_{i} \int_{\Omega} z_{i} c_{i} N_{\mathrm{A}} e \mathrm{d}^{3} \Omega= 0
\end{equation}
where $\sigma_{\mathrm{G}}$ is the charge density in graphene,
$S_{\mathrm{G}}$ is the total area of graphene, and $\Omega$
corresponds to the entire volumetric domain of electrolyte
solution. Note that the graphene surface potential,
$\psi_{\mathrm{G}}$, is not equivalent to $V_{\mathrm{G}}$ due to
change of graphene's work function upon charging, known as the
graphene quantum capacitance effect\cite{Xia_2009} (\Fig{}
\ref{fig:3}a) following:
\begin{equation}
  \label{eq:Vg}
  V_{\mathrm{G}} = \Delta \phi_{\mathrm{G}} + \psi_{\mathrm{G}}
\end{equation}
where
$\Delta \phi_{\mathrm{G}} = \phi_{\mathrm{G}}(V_{\mathrm{g}}) -
\phi_{\mathrm{G}}(V_{\mathrm{g}}=0)$ is the change of graphene’s work
function. The properties of the PG samples fabricated is close to
intrinsic graphene, with charge density at $V_{\mathrm{G}}=0$ measured
to be 1.4$\times$10$^{11}$ \textit{e}$\cdot$cm$^{-2}$, corresponding
to a work function difference of 48 meV from graphene's Dirac point
(details see Supplementary Section \ref{SI-sec:charge-dens}), much
smaller than the $\Delta \phi_{\mathrm{G}}$ induced by gating.
Therefore we adapt the simplification that the charge neutrality point
of graphene coincides with the Dirac point at $V_{\mathrm{G}}$ =
0. The elementary electronic properties of graphene give:
\begin{equation}
  \label{eq:delta-phiG}
  \Delta \phi_{\mathrm{G}} = \mathrm{sign}(\sigma_{\mathrm{G}}) \frac{\hbar v_{\mathrm{F}}}{e}
  \sqrt{\frac{\pi |\sigma_{\mathrm{G}}|}{e}}
\end{equation}
where $\hbar$ is the reduced Planck constant, and
$v_{\mathrm{F}}=1.1\times10^{6}$ m$\cdot$s$^{-1}$ is the Fermi
velocity of graphene (more details find Supplementary Section
\ref{SI-sec:charge-dens}).  Note that the PG film fabricated
experimentally is double-layer turbostratic (randomly aligned)
graphene, in which we assume the
$\Delta \phi_{\mathrm{G}} - \sigma_{\mathrm{G}}$ dependence follows
Eq. \eqref{eq:delta-phiG}. The theory proposed here was solved
self-consistently by discretizing
Eqs. \eqref{eq:pnp}-\eqref{eq:poisson} using the finite-element method
(FEM), in which the graphene surface potential is coupled with
Eqs. \eqref{eq:electro-neutral}-\eqref{eq:delta-phiG} that were solved
simultaneously to reach convergent numerical solutions of
$\sigma_{\mathrm{G}}$ and $\psi_{\mathrm{G}}$ for a given
$V_{\mathrm{G}}$. Clearly, the above model yields symmetric
characteristics for $\sigma_{\mathrm{G}}$ and $\psi_{\mathrm{G}}$ with
respect to $V_{\mathrm{G}}$ due to the symmetric band structure of
graphene (Eq. \eqref{eq:delta-phiG}). However, for graphene
synthesized by chemical vapor deposition (CVD) used in our
experiments, it is well-recognized that the electron traps are
inherently introduced during synthesis and patterning
\cite{Dean_2010}, effectively reducing the charge density and surface
potential on graphene $V_{\mathrm{G}}<0$ (Supplementary \Fig{}
\ref{SI-fig:trap}), which is related to the fact that observed $\xi$
is much lower within the $V_{\mathrm{G}}<0$ regime in \Fig{}
\ref{fig:2}c. With the nonideality in mind, hereafter, we compare the
calculations and experiments in the regime of $V_{\mathrm{G}}>0$. For
example, \Fig{} \ref{fig:3}c presents the calculated
$\sigma_{\mathrm{G}}$ and $\psi_{\mathrm{G}}$ as a function of
$V_{\mathrm{G}}$ considering the ionic diffusion through a single 20
nm-diameter nanopore drilled on a sheet of semi-infinite double layer
graphene that separates HCR containing KCl solution at $c_{0}$ = 0.1
mM, and LCR at $c_{0}/10$ in axisymmetric coordinates (details see
Supplementary Section \ref{SI-fig:scheme}). Indeed, the interplay
between graphene quantum capacitance and the EDL significantly reduces
the graphene surface potential $\psi_{\mathrm{G}}$ to $\sim$0.3 V at
$V_{\mathrm{G}}$ = 1.25 V, corresponding a surface charge density
$\sigma_{\mathrm{G}}$ of $\sim$0.08 C$\cdot$m$^{-2}$. Note that as we
mentioned earlier, an important merit of the 2D nanopore system
considered here is that the surface charge density can be precisely
determined and controlled, rather than being treated as a fitting
parameter as in most of the 2D nanopore literature
(e.g. Ref. \citenum{Rollings_2016}). In addition, we anticipate that
the gate-controlled ionic transport can also be achieved by replacing
graphene with other nanoporous conducting 2D materials, although the
operational gate voltage window required to achieve the same degree of
salt rejection may be narrower, because graphene has the smallest
quantum capacitance (near the Dirac point) among all known 2D
materials \cite{Tian_2016}.
